Simulating High-Dimensional Multivariate Data

### 2021-02-28 010-background.Rmd
renv+box fully implemented
Addressing TJK edits

To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

>>>>

To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

### 2021-02-28 030-bigsimr-package.Rmd
renv+box fully implemented
Addressing TJK edits

This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package.
`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL.
In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (PROVIDE URL) that interfaces with the Julia `Bigsimr` package. 
The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.

>>>>

This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package (https://github.com/SchisslerGroup/r-bigsimr).
`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package (https://github.com/adknudson/Bigsimr.jl).
In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (https://github.com/SchisslerGroup/python-bigsimr/) that interfaces with the Julia `Bigsimr` package. 
The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.

### 2021-02-28 index.Rmd
Addressing TJK edits

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices. It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

<!-- keywords simulation multivariate models high-dimensional data R packages -->

```{r indexSetup, include=FALSE, cache=FALSE}
set.seed(09212020)
```

@@

<!-- 
Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-->

Another central issue lies in characterizing dependency between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
We do, however, provide Pearson matching with some caveats.

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tranrmarkdown::render_site(encoding = 'UTF-8')
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices. It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Monte carlo studies quantify the accuracy and scalability of our approach, up to $d=20,000$. Finally, we demonstrate example applications enabled via `bigsimr` by applying these methods to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

*Keywords:* High-dimensional data, Big-data computing, Simulation, Nonparametric correlation, Gaussian copula, RNA-sequencing data

```{r indexSetup, include=FALSE, cache=FALSE}
library(tidyverse)
set.seed(09212020)
```

@@

<!-- 
Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-->
Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
We do, however, provide Pearson matching with some caveats.
